\Large{ Authors' reply to the comments by Anonymous Referee (Referee \#2) of the manuscript gmd-2017-119} \\ \LARGE \textbf{Plume-SPH 1.0: A three-dimensional, dusty-gas volcanic plume model based on smoothed particle hydrodynamics}

### 2018-03-19 Manuscripts_GMD_ap_ref2.tex
cleanups

\subsection{Parallellism and performance}
One disadvantage of the 3D model is that it usually takes a much longer time than 1D models to complete one simulation. This disadvantage further prevents doing simulation with finer resolution and accounting for more physics in one model. Non-intrusive uncertainty analysis, which is commonly adopted in hazard forecasting, requires finishing multiple simulations within a given time window. High performance computing is therefore essential. Among existing CPU parallel SPH schemes, most of them focus on neighbor search algorithm and dynamic load balancing \citep {ferrari2009new, crespo2015dualsphysics}. Less attention has been paid to developing more flexible data management schemes for more complicated problems. Motivated by techniques developed for mesh based methods, we develop a complete framework for parallelizing SPH with distrubted memory parallelism (MPI) allowing flexible and efficient data access.

The time complexity for raw SPH method is $O(NN)$. Where, $N$ is the total number of SPH particles. Efficient neighbour searching and compact-supported kernel function could help to reduce computational cost. Of the many possible choices we adopt a background grid which was proposed by \citet {monaghan1985refined} and is quite popular in parallel SPH. Then the time complexity is reduced to $O(MN+mN)$. Where $m$ is the average of number of particles within the compact support of kernel function, $M$ is average number of particles in sub-domain within which neighbour searching is carried out. This background grid is also used for domain decomposition. We refer to the elements of background grid, namely squares for two dimensions and cubic for three dimensions, as buckets. 
As for the actual storage of the physical quantities associated to each particle, different strategies have been adopted in existing implementations of SPH.

In DualSPHysics \citep {crespo2015dualsphysics}, the physical quantities of each particle (position, velocity, density...) are stored in arrays. The particles (and the arrays with particle data) are reordered following the order of the cells. This has two advantages: 1) access pattern is more regular and more efficient, 2) it is easy to identify the particles that belong to a cell by using a range since the first particle of each cell is known. But adding, deleting and especially accessing of particles are cumbersome. \citet {ferrari2009new} adopted linked lists using pointers so that particles can be deleted or added during the simulation. Storage problems caused by fixed-size arrays are thereby also eliminated. We define C++ classes which contain all data of particles and buckets. As for the management of data, we adopt hash tables to store pointers to particles and buckets, which gives us not only flexibility of deleting and adding of elements, but also quicker access compared with linked lists. Instead of using the "nature manner" to number particles, we adopt a space filling curve (SFC) based index to give each particle and background bucket an unique identifier -- a strategy known to preserve data locality at minimal cost. The SFC based numbering strategy is further extended to include time step information so that particles added at the same position but different time have different identifiers.

>>>>

\subsection{Parallellism and performance}
One disadvantage of the 3D model is that it usually takes a much longer time than 1D models to complete one simulation. This disadvantage further prevents doing simulation with finer resolution and accounting for more physics in one model. Non-intrusive uncertainty analysis, which is commonly adopted in hazard forecasting, requires finishing multiple simulations within a given time window. High performance computing is therefore essential. Among existing CPU parallel SPH schemes, most of them focus on neighbor search algorithm and dynamic load balancing \citep {ferrari2009new, crespo2015dualsphysics}. Less attention has been paid to developing more flexible data management schemes for more complicated problems. Motivated by techniques developed for mesh based methods, we develop a complete framework for parallelizing SPH with distriubted memory parallelism (MPI) allowing flexible and efficient data access.

The time complexity for an SPH method is $O(N^2)$, where, $N$ is the total number of SPH particles. Efficient neighbor searches and compact-supported kernel functions can help to reduce computational cost. Of the many possible choices we adopt a background grid  which was proposed by \citet {monaghan1985refined} and is quite popular in parallel SPH. Then the time complexity is reduced to $O(MN+mN)$, where $m$ is the average of number of particles within the compact support of the kernel function, $M$ is the average number of particles in sub-domain within which neighbor searches are carried out. This background grid is also used for domain decomposition. We refer to the elements of background grid, namely squares for two dimensions and cubic for three dimensions, as buckets. 
As for the actual storage of the physical quantities  different strategies have been adopted in existing implementations of SPH.

In DualSPHysics \citep {crespo2015dualsphysics}, the physical quantities of each particle (position, velocity, density...) are stored in arrays. The particles (and the arrays with particle data) are reordered following the order of the cells. This has two advantages: 1) access pattern is more regular and more efficient, 2) it is easy to identify the particles that belong to a cell by using a range since the first particle of each cell is known. But adding, deleting and especially accessing of particles are cumbersome. \citet {ferrari2009new} adopted linked lists using pointers so that particles can be deleted or added during the simulation. Storage problems caused by fixed-size arrays are thereby also eliminated. We define C++ classes which contain all data of particles and buckets. As for the management of data, we adopt hash tables to store pointers to particles and buckets, which gives us not only flexibility of deleting and adding of elements, but also quicker access compared with linked lists. Instead of using the "natural order" to number particles, we adopt a space filling curve (SFC) based index to give each particle and background bucket an unique identifier -- a strategy known to preserve data locality at minimal cost. The SFC based numbering strategy is further extended to include time step information so that particles added at the same position but different time have different identifiers.

### 2018-03-19 Reply-to-RC2.tex
cleanups

\documentclass[10pt,a4paper]{article}
%\documentclass[10pt,a4paper]{beamer}
%\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\usepackage{kpfonts}
\usepackage[numbers]{natbib}
%other options include: round, curly, angle, or default options

@@

\maketitle
The comments by the reviewer are recited in italics, followed by our reply in upright font. Equation, section, and figure numbers that  in our response are corresponding to these in the original 
version of GMD discussion paper. \\[12pt]

@@

Thanks for pointing out the most recent progress in volcanic plume modeling. 
Work by \citet{costa2018understanding} is a strong backup for the necessarity of developing more comprehensive 3D computational model. We also came up with some application ideas for Plume-SPH after reading this paper. Citation of this paper is added in "Introduction" section to emphasize the advantage of developing more comprehensive plume model based on SPH.

We have been considering coupling volcanic plume model with magma reservoir model. It is critical for any plume simulation to use more accurate eruption conditions, which might be obtained from magma reservoir model. Thanks for pointing out a new degassing model to infer magma dynamics in volcanic plumes \cite{terray2018new}. Comments regarding coupling volcanic plume model with magma reservoir model is added in conclusion section of this paper along with proper citation.
\\[3pt]

@@

My understanding on what does the reviewer mean by "theoretical bound for computational effort" is the time complexity. Apologize if we misunderstood that term.

Without accounting for detailed algorithm, the time complexity for raw SPH method is $O(NN)$. Where, $N$ is the total number of particles (SPH particles are essentially discretization points and equivalent to grids in mesh-based methods). By adopting background mesh and compact-supported kernel function, the time complexity will be reduced to $O(MN)$. Where $M$ is the average of number of particles within the compact support of kernel function. Please be noted that number of particles within the compact support of kernel function is not constant in our case.

When computations are performed across different processors, theoretically speaking, the total computational time will be $(s+\frac{p}{n}+\bar{p})$. Where $(s+p)\sim O(MN)$ is sequential computational time with $p$ representing computation that parallelizable and $s$ representing computation that can not be parallelized. $n$ is total number of processors. $\bar{p}$ is extra computation introduced by parallelization, for example, 
unecessary duplicating work, communication overhead, time to split and combine.

It requires analyzing on detailed algorithms to obtain more accurate theorical bound. We admit that such theoretical time complexity would be very helpful for future performance optimiztion. However, considering HPC is not the major focus of this paper, we decided to reserve this topic for future research. We thanks reviewer again for proposing this idea.\\[3pt]

\textit{It is not clear for me how the parallelization is performed, for instance, for a given time, do you split the domain across different processors? in such case, what constraints must be satisfied at each local domain in order to guaranty a consistent numerical solution of your equations?. Other possibility is to speed-up matrix computations, is this your case? or both?}

The parallelization is not covered with enough details in this paper as it has been addressed in a separate paper \cite {cao2017data}. Sorry for confusing you. The parallelization is achieved only by splitting computational domain (Fig. 2(c) in GMD discussion paper shows a typical domain decomposition). No matrix computation is involved in SPH scheme, so there is no parallelization regarding solving matrix in our case.
For any subdomain, information from its neighbouring subdomains is required when updating physical quantities.   To guarantee consistency, data is synchronized after each updating of physical quantity.

To address your questions and clean up confusions, we made major revision in section 3.10 (is section 3.9 in the revised manuscript). In addition, portion of our response to reviewer's second comments is also added in section 3.10.

We thanks reviewer again for the careful reading and constructive comments.\\[12pt]

@@

Here is a summary on major adjustments made in the revised manuscript: 
Major revision is made in "Parallellism and performance" section (sectiom 3.10 in original manuscript and 3.9 in the revised manuscript).

\bibliographystyle{plainnat}
\bibliography{Reference}
\end{document}

>>>>

\documentclass{article}
%\documentclass[10pt,a4paper]{beamer}
%\usepackage[utf8]{inputenc}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage[left=2cm,right=2cm,top=2cm,bottom=2cm]{geometry}
\usepackage{kpfonts}
\usepackage[numbers]{natbib}
%other options include: round, curly, angle, or default options

@@

\maketitle
The comments by the reviewer are recited in italics, followed by our reply in upright font. Equation, section, and figure numbers   in our response are corresponding to these in the original 
version of GMD discussion paper. \\[12pt]

@@

Thanks for pointing out some of the more recent work on volcanic plume modeling. 
Work by \citet{costa2018understanding} motivates the development of more comprehensive 3D computational models like those in this paper. Reading this paper has inspired new application ideas for Plume-SPH. Citation of this paper is added in "Introduction" section to emphasize the advantage of developing more comprehensive plume model based on SPH.

We have considered coupling a volcanic plume model with a magma reservoir model. It is critical for any plume simulation to use more accurate eruption conditions, which might be obtained from a magma reservoir model. Thank you for pointing out the new degassing model to infer magma dynamics in volcanic plumes \cite{terray2018new}. Comments regarding coupling volcanic plume model with magma reservoir model are added in conclusion section of this paper along with proper citation.
\\[3pt]

@@

Our interpretation of  the reviewer's request for "theoretical bound for computational effort" is the operational complexity of the algorithms used. Apologize if we  have misunderstood .

Without accounting for the detailed algorithm, the complexity for raw SPH method is $O(N*N)$ operations. Where, $N$ is the total number of particles (SPH particles are essentially equivalent to discretization points  in mesh-based methods). By adopting a background mesh and a compact-supported kernel function, the  complexity will be reduced to $O(MN+mN)$. Where $m$ is the average of number of particles within the compact support of the kernel function, $M$ is number of particles among which neighbor searches are carried out. Please note that number of particles within the compact support of the kernel function is not constant in our case.

When computations are performed across different processors, theoretically speaking, the total computational time will be $(s+\frac{p}{n}+\bar{p})$,  where $(s+p)$(time needed for$ \sim O(MN+mN)$ operations) is total sequential computational time with $p$ representing computation that is parallelizable and $s$ represents computation that can not be parallelized,  $n$ is total number of processors and $\bar{p}$ is extra computation and communication introduced by parallelization, for example, 
unnecessary duplication of work, communication overhead, time to split and combine.

It requires complete analysis of the detailed algorithms and instrumented data gathering to obtain  more accurate theoretical and achieved bounds. We agree that such  time complexity analysis would be very helpful for future performance optimization. However, considering HPC is not the major focus of this paper, we would like to reserve this topic for future research. We thank the reviewer again for proposing this idea.\\[3pt]

\textit{It is not clear for me how the parallelization is performed, for instance, for a given time, do you split the domain across different processors? in such case, what constraints must be satisfied at each local domain in order to guarantee a consistent numerical solution of your equations?. Other possibility is to speed-up matrix computations, is this your case? or both?}

The parallelization is not covered with enough details in this paper as it has been addressed in a separate paper \cite {cao2017data}.  The parallelization is achieved only by splitting the computational domain (Fig. 2(c) in GMD discussion paper shows a typical domain decomposition). No matrix computation is involved in the SPH scheme, so there is no parallelization regarding solving the matrix in our case.
For any subdomain, information from its neighboring subdomains is required when updating physical quantities.   To guarantee consistency, data is synchronized after each updates of physical quantities in the shared (or "halo") regions.

To address your questions and clean up confusion, we made a major revision in section 3.10 (is section 3.9 in the revised manuscript). In addition, portion of our response to reviewer's second comments are also added in section 3.10.

We thank reviewer again for the careful reading and constructive comments.\\[12pt]

@@

Here is a summary on major adjustments made in the revised manuscript: 
Major revision is made in "Parallelism and performance" section (section 3.10 in original manuscript and 3.9 in the revised manuscript). 
\begin{thebibliography}{3}
\providecommand{\natexlab}[1]{#1}
\providecommand{\url}[1]{\texttt{#1}}
\expandafter\ifx\csname urlstyle\endcsname\relax
  \providecommand{\doi}[1]{doi: #1}\else
  \providecommand{\doi}{doi: \begingroup \urlstyle{rm}\Url}\fi

\bibitem[Cao et~al.(2017)Cao, Patra, and Jones]{cao2017data}
Zhixuan Cao, Abani Patra, and Matthew Jones.
\newblock Data management and volcano plume simulation with parallel sph method
  and dynamic halo domains.
\newblock \emph{Procedia Comput. Sci.,}, 108:\penalty0 786--795, 2017.

\bibitem[Costa et~al.(2018)Costa, Suzuki, and
  Koyaguchi]{costa2018understanding}
Antonio Costa, Yujiro Suzuki, and Takehiro Koyaguchi.
\newblock Understanding the plume dynamics of explosive super-eruptions.
\newblock \emph{Nature communications}, 9\penalty0 (1):\penalty0 654, 2018.

\bibitem[Terray et~al.(2018)Terray, Gauthier, Salerno, Caltabiano, Spina,
  Sellitto, and Briole]{terray2018new}
Luca Terray, Pierre-J Gauthier, Giuseppe Salerno, Tommaso Caltabiano,
  Alessandro~La Spina, Pasquale Sellitto, and Pierre Briole.
\newblock A new degassing model to infer magma dynamics from radioactive
  disequilibria in volcanic plumes.
\newblock \emph{Geosciences}, 8\penalty0 (1):\penalty0 27, 2018.

\end{thebibliography}
%\bibliographystyle{plainnat}
%\bibliography{Reference}
\end{document}
